Lactate transport inhibition therapeutically reprograms fibroblast

### 2024-03-03 manuscript/manuscript.qmd
finish results draft and remake all targets

## Novel MCT4 inhibitor improves experimental pulmonary fibrosis

Our preclinical data suggest that MCT4 inhibition, as a single drug therapy, is more potently antifibrotic than MCT1 inhibition, both in terms of inhibiting myofibroblast differentiation and decreasing bleomycin-induced fibrosis. VB124

>>>>

## Novel MCT4 inhibitor mitigates experimental pulmonary fibrosis

Our preclinical data suggest that MCT4 inhibition, as a single target therapy, is more potently antifibrotic than MCT1 inhibition, both in terms of inhibiting myofibroblast differentiation *in vitro* and decreasing bleomycin-induced fibrosis. VB253 is a novel inhibitor of MCT4 with similar selectivity, but ~10-fold increased potency, for MCT4 inhibition. Like VB124, VB253 dose-dependently decreases **TGFβ-stimulated?** α-SMA in human IPF lung fibroblasts *ex vivo* (**Figure 10A**). Given the robust effect of MCT4 inhibition on myofibroblast differentiation, we compared VB253 to nintedanib, an anti-fibrotic medication clinically indicated for pulmonary fibrosis (**Figure 10A-B**). Nintedanib inhibits TGFβ-mediated fibroblast to myofibroblast differentiation and decreases Col1a1 expression *in vitro* and *in vivo*  [@wollin2014]. While both drugs potently attenuated α-SMA expression, nintedanib demonstrated a moderate cytotoxic effect that was not observed with VB253 (**Figure 10B**). Nintedanib prevents myofibroblast differentiation, in part, through inhibition of TGFβ receptor phosphorylation and activation of Smad-dependent signaling pathways [@rangarajan2016]. Consistent with these findings, nintedanib dose-dependently attenuates TGFβ-mediated Smad3 phosphorylation, while VB253 and VB124 have no impact on this signaling pathway (**Figure 6**, **Supplemental Figure 10**).

We next assessed the ability of VB253 to inhibit bleomycin-induced pulmonary fibrosis. In experiments conducted independently from those reported above, VB253 was administered to young mice (8-10 weeks) beginning on day 7 following bleomycin (**Figure 10C**). Mice treated with nintedanib or pirfenidone provided comparisons to currently available therapies for pulmonary fibrosis. Three weeks after bleomycin administration, whole-body plethysmography was performed to assess breathing patterns in unrestrained mice. Enhanced pause (Penh) is a dimensionless index describing airflow during tidal breathing found to increase significantly in bleomycin-treated mice [@vanoirbeek2010]. VB253 restored Penh to baseline levels, suggesting normalization of respiratory patterens in bleomycin-treated mice (**Figure 10D**). This was associated with decreased pulmonary fibrosis by histologic assessment (**Figure 10E**) and α-SMA expression (**Figure 10F-G**). Morevover, the effects of VB253 compared favorably to the antifibrotic effects of nintedanib and pirfenidone.

Compared to young mice, bleomycin-induced fibrosis in aged mice is more severe and does not resolve spontaneously [@redente2011]. Given that IPF incidence increases with age, some studies have suggested such aged mouse models may be more clinically relevant. Thus, we examined the antifibrotic effects of VB253 in aged (60+ weeks) mice. Similar to young mice, VB253 decreased fibrosis severity as quantified by Ashcroft score and α-SMA expression (**Figure 10H-J**). The magnitude of improvement was similar to that observed with nintedanib and pirfenidone. As anticipated, VB2253 decreased total lung lactate similar to VB124 (**Figure 9A**). Taken together, these data provide compelling preclinical data for the development of lactate transporter inhibition as a therapeutic strategy for fibrotic lung disease.
